---
title: "Re-emergence of RSV in the United States"
author: "Dan Weinberger"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
#rmarkdown::render('rsv_covid.Rmd', output_file = 'docs/index.html')
knitr::opts_chunk$set(echo = F)
#install.packages('curl')
#library(curl)
#if (!require("devtools")) install.packages("devtools")

#Install dev version!
#devtools::install_github("PMassicotte/gtrendsR")

library(RApiDatetime)
library(gtrendsR)
library(pbapply)
library(reshape2)
library(MMWRweek)
library(dplyr)
library(ggplot2)
library(geofacet)
library(lubridate)
library(plotrix)
source('./R/runIfExpired.R')

updateData = F

```

In these plots, we pulled searches for 'RSV' from Google trends using the gtrendsR package. Previous work found that Google searches for 'RSV' capture spatiotemporal variations in RSV activity across the US: https://www.biorxiv.org/content/10.1101/607119v1


```{r}

if(updateData==T){
  t1=0
}else{
  t1=99999999999999999999
}

state.abb.us <- paste0('US-', state.abb)
searches <- c('rsv')


state.trends <- runIfExpired('state.trends',maxage=t1, ~pblapply(state.abb.us, function(x) {
    lapply(searches, function(y){
        gtrends(keyword=y, geo = c(x), category = 0, gprop='web')
    })
  }
)
)


#Other geographies
#
#data("countries")
#View(countries)
test.grps <- list( c( 'RSV','ZA' ), 
                  c('bronquiolitis', 'AR'), 
                  c('bronquiolite','BR'),
                  c('bronchiolite','FR'),

                  c('bronchiolite','FR-J'),
                  c('RSV','GB')
      )
names(test.grps) <- c('S. Africa','Argentina', 'Brazil','France', 'France-Paris region')

country.trends <- runIfExpired('country.trends',maxage=t1, ~lapply(test.grps, function(x) gtrends(keyword=x[1], geo = x[2], category = 0, gprop='web',onlyInterest=T))
)


#Metro region
#https://stackoverflow.com/questions/61213647/what-do-gtrendsr-statistical-areas-correlate-with
#Nielsen DMA map: http://bl.ocks.org/simzou/6459889
data("countries")

metros <- countries[countries$country_code == 'US', ]
metros <-
  metros[grep("[[:digit:]]", substring(metros$sub_code, first = 4)), ]
metros$numeric.sub.area <- gsub('US-', '', metros$sub_code)

name.spl <- stringr::str_split(metros$name, ' ')

find.state.loc <-
  lapply(name.spl,  function(x)
    which(x %in% state.abb)) #position of state in string

state.metros <-
  mapply(function(x, y) {
    y[x]
  } , x = find.state.loc, y = name.spl)

n.states <- lapply(state.metros, length)


#This provides a full list of US-STATE-METRO codes; including when a metro crosses 2 states. Note that not all of these codes are valid
all.metro.combos <- mapply(function(state, metro) {
  paste('US', state, metro, sep = '-')
},
state.metros,
metros$numeric.sub.area, SIMPLIFY = F)


all.metro.combos <- unlist(all.metro.combos[n.states > 0])
all.metro.combos <- all.metro.combos[!is.na(all.metro.combos)]

searches <- c('rsv')

metro.trends <- runIfExpired('metro.trends',maxage=t1, ~pblapply(all.metro.combos, function(x) {
    lapply(searches, function(y){
        tryCatch(gtrends(keyword=y, geo = c(x), category = 0, gprop='web',onlyInterest=T), error=function(e) NULL)
    })
  }
)
)
names(metro.trends) <- metros$name[n.states > 0]



```



```{bash, eval=F}
python -m pip install --upgrade pip

python -m pip install pytrends
pip install pytrends --upgrade

```

```{python, eval=F}
from pytrends.request import TrendReq  

pytrends = TrendReq(hl='en-US', tz=360)   

kw_list = ['rsv'] # list of keywords to get data   

pytrends.build_payload(kw_list, cat=0)   

data = pytrends.interest_over_time()  

data = data.reset_index()    


```



## US States

Searches for "RSV"
```{r, results=FALSE, message=FALSE, warning=FALSE}
l1a <- state.trends
#l1 <- c(l1a, l1b)
l2 <- lapply(l1a,function(x) lapply(x, '[[','interest_over_time'))
l3 <- lapply(l2, function(x) do.call( 'rbind.data.frame', x))

d1 <- do.call( 'rbind.data.frame', l3)

d1$geo <-gsub('US-','',d1$geo)

d1.m <- melt(d1[,c('date','hits','geo', 'keyword')], id.vars=c('geo','date','keyword'))

d1.m$value <- as.numeric(d1.m$value)
#hist(d1.m$value[d1.m$geo=='US-FL'])
d1.c <- acast(d1.m ,date~geo~keyword)


#Case by weekN and year
d1.m.alt <- d1.m

dates2 <- MMWRweek(as.Date(d1.m$date))
d1.m.alt <- cbind.data.frame(d1.m.alt,dates2[,c('MMWRyear', 'MMWRweek')])
max.wk.yr <-  max(d1.m.alt$MMWRweek[d1.m.alt$MMWRyear==max(d1.m.alt$MMWRyear)] )


d1.m.alt$epiyr <- d1.m.alt$MMWRyear

d1.m.alt$epiyr[d1.m.alt$MMWRweek<=max.wk.yr] <- d1.m.alt$MMWRyear[d1.m.alt$MMWRweek<=max.wk.yr] - 1

d1.c.alt <- acast(d1.m.alt ,MMWRyear~MMWRweek~geo~keyword, fun.aggregate = sum)


 last.wk <- max(d1.m.alt$MMWRweek[d1.m.alt$MMWRyear==max(d1.m.alt$MMWRyear)] )

 lastyr.char <- dimnames(d1.c.alt)[[1]][dim(d1.c.alt)[1]]
 
 d1.c.alt[lastyr.char,as.numeric(dimnames(d1.c.alt)[[2]])>last.wk,,] <- NA


```




```{r, fig.width=8, fig.height=6,results=FALSE, message=FALSE, warning=FALSE}

df1 <- dcast(d1.m.alt[d1.m.alt$keyword=='rsv',] ,MMWRyear+MMWRweek+geo~., fun.aggregate = sum)
names(df1) <- c('year','week','state','cases')

df1$state <- gsub('US-','',df1$state)
cols <- c(rep('gray',4),rgb(1,0,0,0.2),'blue')
ltypes <- c(rep(1,3),2,1,1)

my_us_grid <- us_state_grid1[c(-2, -11,-51), ]

df1$year <- as.factor(df1$year)
theme_set(theme_minimal())
p1 <- ggplot(df1, aes(
  x = week,
  y = cases,
  group = year,
  colour = year
)) +
  geom_line(aes(lty = year)) +
  scale_color_manual(values = cols) +
  scale_linetype_manual(values = ltypes) +
  labs(title = "Searches for 'RSV in the US' (Google trends)",subtitle='By calendar year',
       x = element_blank(),
       y = element_blank()) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    legend.position = "none",
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    plot.title = element_text(size = 24),
    strip.background = element_rect(color = "white")
  ) +
  facet_geo( ~ state, grid = my_us_grid)
p1
```



## Selected states, by calendar year

```{r, fig.width=8, fig.height=6, results=FALSE, message=FALSE, warning=FALSE}
cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue')
linetypes=c(rep(1,5),2,2)
states <- dimnames(d1.c.alt)[[3]]

select.states <-c('FL','GA','NY','CA','WA','MD','TX','IL','WI')
cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue')
linetypes=c(rep(1,5),2)
states <- dimnames(d1.c.alt)[[3]]
par(mfrow=c(3,3), mar=c(2,2,1,0))
lapply(select.states, function(x){
  matplot(t(d1.c.alt[,1:52,x,'rsv']), type='l', col=cols, bty='l', xlab='', ylab='', xlim=c(1,52), main=x, lty=linetypes)
abline(v=36, lty=2, col='gray')
})

```

## Polar plots of states

This plots the time series on a circular axis, showing the shift in timing compared to normal. Blue shows the periods since July 2020

```{r, fig.width=8, fig.height=6,results=FALSE, message=FALSE, warning=FALSE}
df1$year <- as.factor(df1$year)

cols <- c(rep('gray',4),'red','blue')
linetypes=c(rep(1,5),1,1)

theme_set(theme_minimal())
p1 <- ggplot(df1, aes(
  x = week,
  y = cases,
  group = year,
  colour = year
)) +
  geom_line(aes(lty = year)) +
  coord_polar() +
  scale_color_manual(values = cols) +
  scale_linetype_manual(values = ltypes) +
  labs(title = "Searches for 'RSV' in the US (Google trends)",
       x = element_blank(),
       y = element_blank()) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    legend.position = "none",
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    plot.title = element_text(size = 24),
    strip.background = element_rect(color = "white")
  ) +
  facet_geo( ~ state, grid = my_us_grid)
p1

```




## Other Countries, by calendar year


```{r results=FALSE, message=FALSE, warning=FALSE}
l1b <- country.trends

l2 <- lapply(l1b,  '[[','interest_over_time')

d1 <- do.call( 'rbind.data.frame', l2)

d1.m <- melt(d1[,c('date','hits','geo')], id.vars=c('geo','date'))

d1.m$value <- as.numeric(d1.m$value)
#hist(d1.m$value[d1.m$geo=='US-FL'])
d1.c <- acast(d1.m ,date~geo)


#Case by weekN and year
d1.m.alt <- d1.m

dates2 <- MMWRweek(as.Date(d1.m$date))
d1.m.alt <- cbind.data.frame(d1.m.alt,dates2[,c('MMWRyear', 'MMWRweek')])
max.wk.yr <-  max(d1.m.alt$MMWRweek[d1.m.alt$MMWRyear==max(d1.m.alt$MMWRyear)] )

d1.m.alt$epiyr <- d1.m.alt$MMWRyear
d1.m.alt$epiyr[d1.m.alt$MMWRweek<=max.wk.yr] <- d1.m.alt$MMWRyear[d1.m.alt$MMWRweek<=max.wk.yr] - 1
d1.m.alt$epiwk <- d1.m.alt$MMWRweek
d1.m.alt$epiwk[d1.m.alt$MMWRweek<=max.wk.yr] <- d1.m.alt$MMWRweek[d1.m.alt$MMWRweek<=max.wk.yr] + 52
#d1.m.alt$geo <- substr(d1.m.alt$geo,4,5)

d1.c.alt <- acast(d1.m.alt ,MMWRyear~MMWRweek~geo, fun.aggregate = sum)


last.wk <- max(d1.m.alt$MMWRweek[d1.m.alt$MMWRyear==max(d1.m.alt$MMWRyear)] )
lastyr.char <- dimnames(d1.c.alt)[[1]][dim(d1.c.alt)[1]]
d1.c.alt[lastyr.char,as.numeric(dimnames(d1.c.alt)[[2]])>last.wk,] <- NA

countries <- c('GB','FR','FR-J')

cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue')
states <- dimnames(d1.c.alt)[[3]]
par(mfrow=c(1,1), mar=c(2,2,1,0))
lapply(countries, function(x){
  matplot(t(d1.c.alt[,1:52,x]), type='l', col=cols, bty='l', xlab='', ylab='', xlim=c(1,52), main=x, lty=linetypes)

})
```



```{r, fig.width=4, fig.height=3, results=FALSE, message=FALSE, warning=FALSE}
l2 <- lapply(l1b,  '[[','interest_over_time')

d1 <- do.call( 'rbind.data.frame', l2)

d1.m <- melt(d1[,c('date','hits','geo')], id.vars=c('geo','date'))

d1.m$value <- as.numeric(d1.m$value)
#hist(d1.m$value[d1.m$geo=='US-FL'])
d1.c <- acast(d1.m ,date~geo)


#Case by weekN and year
d1.m.alt <- d1.m
dates2 <- MMWRweek(as.Date(d1.m$date))
d1.m.alt <- cbind.data.frame(d1.m.alt,dates2[,c('MMWRyear', 'MMWRweek')])


d1.c.alt <- acast(d1.m.alt ,MMWRyear~MMWRweek~geo, fun.aggregate = sum)

last.wk <- max(d1.m.alt$MMWRweek[d1.m.alt$MMWRyear==max(d1.m.alt$MMWRyear)] )
d1.c.alt[dimnames(d1.c.alt)[[1]][dim(d1.c.alt)[1]]
,as.numeric(dimnames(d1.c.alt)[[2]])>last.wk,] <- NA

countries <- c( 'AR','BR','ZA')

cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue')

linestypes <- c(rep(1,5),2)

states <- dimnames(d1.c.alt)[[3]]
par(mfrow=c(1,1), mar=c(2,2,1,0))
lapply(countries, function(x){
  matplot(t(d1.c.alt[,,x]), type='l', col=cols, bty='l', xlab='', ylab='', xlim=c(1,52), main=x, lty=linestypes)

})

```


## US Counties

```{r}
l1a <- metro.trends
#l1 <- c(l1a, l1b)
l2 <- lapply(l1a,function(x) lapply(x, '[[','interest_over_time'))
l3 <- lapply(l2, function(x) do.call( 'rbind.data.frame', x))

for(i in 1:length(l3)){
  if(nrow(l3[[i]])>0 ){
   l3[[i]]$city_name= names(l1a)[i]
  }
}


d1 <- do.call( 'rbind.data.frame', l3)

d1.m <- melt(d1[,c('date','hits','city_name', 'keyword','geo')], id.vars=c('geo','city_name','date','keyword'))

d1.m$value <- as.numeric(d1.m$value)
#hist(d1.m$value[d1.m$geo=='US-FL'])
d1.c <- acast(d1.m ,date~geo+city_name~keyword)


#Case by weekN and year
d1.m.alt <- d1.m

dates2 <- MMWRweek(as.Date(d1.m$date))
d1.m.alt <- cbind.data.frame(d1.m.alt,dates2[,c('MMWRyear', 'MMWRweek')])
max.wk.yr <-  max(d1.m.alt$MMWRweek[d1.m.alt$MMWRyear==max(d1.m.alt$MMWRyear)] )

d1.m.alt$epiyr <- d1.m.alt$MMWRyear
d1.m.alt$epiyr[d1.m.alt$MMWRweek<=max.wk.yr] <- d1.m.alt$MMWRyear[d1.m.alt$MMWRweek<=max.wk.yr] - 1

d1.m.alt$epiwk <- d1.m.alt$MMWRweek
d1.m.alt$epiwk[d1.m.alt$MMWRweek<=max.wk.yr] <- d1.m.alt$MMWRweek[d1.m.alt$MMWRweek<=max.wk.yr] + 52
#d1.m.alt$geo <- substr(d1.m.alt$geo,4,5)

d1.c.alt <- acast(d1.m.alt ,MMWRyear~MMWRweek~geo+city_name~keyword, fun.aggregate = sum)


last.wk <- max(d1.m.alt$MMWRweek[d1.m.alt$MMWRyear==max(d1.m.alt$MMWRyear)] )
lastyr.char <- dimnames(d1.c.alt)[[1]][dim(d1.c.alt)[1]]
d1.c.alt[lastyr.char,as.numeric(dimnames(d1.c.alt)[[2]])>last.wk,,] <- NA
```

## Selected counties

Florida

```{r, fig.width=10, fig.height=10, results=FALSE, message=FALSE, warning=FALSE}
cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue')
linetypes=c(rep(1,5),2,2)
select.counties <- grep('US-FL', dimnames(d1.c.alt)[[3]] )

select.counties2 <- c(grep('US-MA', dimnames(d1.c.alt)[[3]] ),
grep('US-CT', dimnames(d1.c.alt)[[3]] ),
grep('US-NH', dimnames(d1.c.alt)[[3]] ),
grep('US-RI', dimnames(d1.c.alt)[[3]] ),
grep('US-VT', dimnames(d1.c.alt)[[3]] ))

county.labs <- substring( dimnames(d1.c.alt)[[3]], first=11)

cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue')
linetypes=c(rep(1,5),2)
counties <- dimnames(d1.c.alt)[[3]]
```

Florida

```{r, fig.width=10, fig.height=10, results=FALSE, message=FALSE, warning=FALSE}
par(mfrow=c(3,3), mar=c(2,2,1,0))
lapply(select.counties, function(x){
  matplot(t(d1.c.alt[,1:52,x,'rsv']), type='l', col=cols, bty='l', xlab='', ylab='', xlim=c(1,52), main=county.labs[x], lty=linetypes)
abline(v=36, lty=2, col='gray')
})
```

New England

```{r, fig.width=10, fig.height=10, results=FALSE, message=FALSE, warning=FALSE}

par(mfrow=c(2,2), mar=c(2,2,1,0))

lapply(select.counties2, function(x){
  matplot(t(d1.c.alt[,1:52,x,'rsv']), type='l', col=cols, bty='l', xlab='', ylab='', xlim=c(1,52), main=county.labs[x], lty=linetypes)
abline(v=36, lty=2, col='gray')
})

```

New England-most recent 52 weeks

```{r, fig.width=10, fig.height=10, results=FALSE, message=FALSE, warning=FALSE}
cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue','blue')
linetypes=c(rep(1,5),2,2)
select.counties <- c(grep('US-CT', dimnames(d1.c.alt)[[3]] ) ,grep('US-RI', dimnames(d1.c.alt)[[3]] ),
                     grep('US-MA', dimnames(d1.c.alt)[[3]] ) ,grep('US-NH', dimnames(d1.c.alt)[[3]] ) ,
                     grep('US-VT', dimnames(d1.c.alt)[[3]] ) ,grep('US-ME', dimnames(d1.c.alt)[[3]] ) ) 
county.labs <- substring( dimnames(d1.c.alt)[[3]], first=11)

cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue')
linetypes=c(rep(1,5),2)
counties <- dimnames(d1.c.alt)[[3]]
par(mfrow=c(3,3), mar=c(2,2,1,0))
lapply(select.counties, function(x){
  matplot(t(d1.c.alt[,1:52,x,'rsv']), type='l', col=cols, bty='l', xlab='', ylab='', xlim=c(1,52), main=county.labs[x], lty=linetypes)
abline(v=36, lty=2, col='gray')
})

```


```{r, fig.width=10, fig.height=10, results=FALSE, message=FALSE, warning=FALSE}
cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue','blue')
linetypes=c(rep(1,5),2,2)
select.counties <- c(grep('US-NY', dimnames(d1.c.alt)[[3]] ) ,grep('US-NJ', dimnames(d1.c.alt)[[3]] ),
                     grep('US-PA', dimnames(d1.c.alt)[[3]] )   ) 
county.labs <- substring( dimnames(d1.c.alt)[[3]], first=11)

cols <- c(rep('gray',4),rgb(1,0,0,0.3),'blue')
linetypes=c(rep(1,5),2)
counties <- dimnames(d1.c.alt)[[3]]
par(mfrow=c(3,3), mar=c(2,2,1,0))
lapply(select.counties, function(x){
  matplot(t(d1.c.alt[,,x,'rsv']), type='l', col=cols, bty='l', xlab='', ylab='', xlim=c(1,52), main=county.labs[x], lty=linetypes)
abline(v=36, lty=2, col='gray')
})

```
